---
title: "Estimate-and-project-trends-using-posterior-distribution-draws"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Estimate-and-project-trends-using-posterior-distribution-draws}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE, 
  message = FALSE,
  comment = "#>"
)
```

This vignette provides an alternate method for estimating and predicting trends, similar to those described in the base workflow with [annual indices data input](https://ninoxconsulting.github.io/birdtrends/articles/Estimate-and-project-trends-using-annual-indices-datasets.html). 

Rather than annual indices, this workflow uses draws of the posterior distributions of the smooth component of population trajectory, drawn from a GAMYE Breeding Bird Survey (BBS) model (using bbsBayes2 package). 

Note: this vignette assumes you had already followed the steps outlined [here](https://ninoxconsulting.github.io/birdtrends/articles/Getting_started.html)

## 1. Set-up

Lets start by loading all the libraries required.

```{r setup, warning = FALSE, message =FALSE}
library(birdtrends)
library(ggplot2)
library(mgcv)
library(dplyr)
library(tidyr)
```

## 2. Extract the smooth component of bbsBayes2 fit

An example data set, draws of the posterior distribution of annual indices is provided within this package of the Pacific Wren ("Troglodytes pacificus"), generated using the [bbsBayes2](https://bbsbayes.github.io/bbsBayes2/) package. This data was kindly provided by A.C. Smith. 
While this dataset was derived using the bbsBayes model, any Bayesdian trend model output will provide suitable input. 


```{r view input dataset}

head(posterior_draws_data)

```

In this example data set we have draws from the posterior ditribution of the trend model from 1968 to 2022.


### 2.1 Fit a generalized additive model to each draw (GAM)

We can fit a General Additive Model (GAM) to each draw to estimate the overall trend for the species over all years, or a specific date range. This model fits a smooth time-series function (i.e., the GAM) to the log-transformed annual estimates of relative abundance.


```{r, fit gam}
indat2 <- as.data.frame(posterior_draws_data)
    
fitted_data <- fit_gam(indat2)
    
```


### 3. Calculate trend 


```{r reformat fitted data from hgam}

trend_sm <- get_trend(fitted_data, start_yr = 2014, end_yr = 2022, method = "gmean")
         
```


We can summarise the trend estimates to provide a median and confidence internal

```{r}
trend_sm |> 
  dplyr::mutate(trend_q0.025 = quantile(trend_log, 0.025),
         trend_q0.500 = quantile(trend_log,0.500),
         trend_q0.975 = quantile(trend_log,0.975)) |> 
  dplyr::select(c(trend_q0.025, trend_q0.500, trend_q0.975)) |> 
  distinct()

```

### 4. Project trend

We can now use our modeled annual indices and estimated trends for our given years to project into the future. 

```{r}

preds_sm <- proj_trend(fitted_data, trend_sm, start_yr = 2023, proj_yr = 2050)
     

```


### 5. Plot the projected values

Now lets plot the results, to make a "pretty plot" we will use all the steps we worked through above. This includes 1) raw observed indices, 2) modeled indices, 3) projected indices generated from our trends. 

Note in the plot below we only had the posterior draws input data and not the raw annual estimates. In these cases where no raw indices is supplied, values are based on 95% of distribution of data. 


```{r}
gam_plot <- plot_trend(raw_indices = NULL,
                             model_indices = fitted_data,
                             pred_indices = preds_sm,
                             start_yr = 2014,
                             end_yr = 2022)

gam_plot

```
