---
title: "Getting_started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting_started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The birdtrends package enables users to estimate trends of populations into the future based on a variety of input data types. 

# Getting set-up 

1) Install birdtrends

```{r, eval=FALSE}
remotes::install_github("ninoxconsulting/birdtrends")
```
2) Set up modelling connections. 

The birdtrends package uses the cmdstanr package to run the hierarchical GAM models using [Stan](https://mc-stan.org/). We can install cmdstanr with:

```{r, eval=FALSE}
install.packages("cmdstanr", repos = c("https://mc-stan.org/r-packages/",
                                       getOption("repos")))
```

The cmdstanr package in turn requires the `cmdstan` program to run Stan programs. You can use the `cmdstanr` package to `cmdstan`:

```{r, eval=FALSE}
cmdstanr::install_cmdstan()
```

And then check that `cmdstan` was installed properly:

```{r, eval=FALSE}
cmdstanr::check_cmdstan_toolchain()
```

# What dataset can be used? 



# what type of models can we use? 


# How to estimate trends 


# Output summaries 


