---
title: "Use annual indices and uncertaintly to estimate trends over time"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Use Annual indices and uncertaintly to estimate trends over time}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  #eval = FALSE,
  collapse = TRUE,
  comment = "#>"
)
```


We can use the `birdtrends` package to model trends overtime based on annual indices datasets. These include information on year of data, index, and some form of uncertainty. 

Note: this vignette assumes you had already followed the steps outlined [here](https://ninoxconsulting.github.io/birdtrends/articles/Getting_started.html)

## 1. Set-up

Lets start by loading all the libraries required.

```{r setup, warning = FALSE, message =FALSE}
devtools::load_all()
library(birdtrends)
library(ggplot2)
library(mgcv)
library(dplyr)
library(tidyr)

```


## 2. Model the annual indices based on input data of observed annual indices. 

In many cases, we may not have access to the breadth of information retained in estimating the original modeled annual indices. 

An example data set is provided within this package. This is an annual estimate on the Pacific Wren (" "), generated using the [bbsBayes2](https://bbsbayes.github.io/bbsBayes2/) package. This data was kindly provided by A.C. Smith. 

```{r}

head(annual_indicies_data)

```

In this example data set we have an annual index from 1968 to 2022, along with confidence intervals of 2.5% and 97.5%. 

### 2.1 Fit a Bayesian Heirachial GAM (HGAM)

We can fit a hierarchical Bayesian General Additive Model (HGAM) to estimate the overall trend for the species over all years, or a specific date range. 


```{r run_hgam, warning = FALSE, message =FALSE}

indat1 <- annual_indicies_data

# fit the Heirachial GAM model 
fitted_data <- fit_hgam(indat1, start_yr = NA, end_yr = NA, n_knots = 5)

```

### 2.2 Explore the results

Select a subset of all the fitted HGAM models and reformat to long format. 
In this example we selected 100 rows for simplicity. 


```{r}
sel_hgams <- fitted_data %>%
  dplyr::slice_sample(., n = 100) %>%  
  dplyr::mutate(draw = seq(1, 100, 1)) %>% 
  tidyr::pivot_longer(., cols = !starts_with("d")) |> 
  dplyr::mutate(yearn = as.integer(name) - min(as.integer(name)))
         

```

We can reformat our original data to use in the plot 

```{r}
indat1 <- indat1 |> 
  dplyr::mutate(yearn = as.integer(year) - min(as.integer(year)))  

```

Now we can plot the predicted posterior distributions of 100 draws, along with the raw annual indices data provided. 

```{r}
comp_plot <- ggplot2::ggplot(data = sel_hgams,
                    ggplot2::aes(x = yearn,y = value,
                        group = draw, colour = draw))+
  ggplot2::geom_pointrange(data = indat1,
                  ggplot2::aes(x = yearn, y = index,
                      ymin = index_q_0.025,
                      ymax = index_q_0.975),
                  inherit.aes = FALSE,
                  alpha = 0.3)+
  ggplot2::geom_line(alpha = 0.3)+
  ggplot2::scale_colour_viridis_c() +
  ggplot2::scale_y_continuous(trans = "log10")+
  ggplot2::theme_bw()

```

```{r, fig.width=5, fig.height= 5}
comp_plot

```


## 3. Estimate trends

We can use our modeled values to estimate a trend over a given period of time. This can be used to predict into the future. 

First we need to reformat our fitted data. We can modify the object 'sel_hgams' above or reformat the raw output. 

We will convert our data to long form, with the following columns "draw", "year", and "proj_y". Note if using the output from the "fit_hgam" we can use the code below. 

```{r}
  ldf <- tibble::rowid_to_column(fitted_data, "draw") %>%
    tidyr::pivot_longer(., cols = !starts_with("d")) %>%
    dplyr::rename('year' = name, "proj_y" = value)%>%
    mutate(year = as.integer(year))

```
This creates a dataset in which each row contains an estimated indices per year and per posterior draw (n = 4000). 

```{r}
 head(ldf)
```


Now we can estimate a trend based on a given time internal and method of estimating trends. 
Where no dates are specified, the minimum and maximum years will be used. Two methods are available to estimate trends; 1) Geometric mean and 2) Linear regression. 


```{r}
  trend_sm <- get_trend(ldf, start_yr = 2014, end_yr = 2022, method = "gmean")

```

```{r}
 head(trend_sm)
```

We can summarise the trend estimates to provide a median and confidence internal

```{r}
trend_sm |> 
  dplyr::mutate(trend_q0.025 = quantile(trend_log, 0.025),
         trend_q0.500 = quantile(trend_log,0.500),
         trend_q0.975 = quantile(trend_log,0.975)) |> 
  dplyr::select(c(trend_q0.025, trend_q0.500, trend_q0.975)) |> 
  distinct()

```

