---
title: "Use Annual indices and uncertaintly to estimate trends over time"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Use Annual indices and uncertaintly to estimate trends over time}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(birdtrends)
```

We can use the `birdtrends` package to model trends overtime based on annual indices datasets. These include information on year of data, index, and some form of uncertainty. 


This vignette assumes you had already followed the steps outlined [here](https://github.com/ninoxconsulting/birdtrends/blob/main/vignettes/Getting_started.Rmd)


## Modelling trends based on annual indices. 

In many cases, we may not have access to the breadth of information retained in estimating the original modeled annual indices. 




An example dataset is provided within this package. This is an annual estimate on the Pacific Wren (")

```{r}
head(annual_indicies_data)

```


